Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-06 ms.tex

Type Ia SNe represent a source of heating, which unlike core collapse SNe is present even in an evolved stellar population.  If each SN Ia injects thermal energy $E_{\rm Ia}$ into the interstellar medium, and the SN rate per stellar mass is given by $R_{\rm   Ia}$, then the resulting volumetric heating rate of $E_{\rm Ia}R_{\rm  Ia}$ produces an effective heating parameter (eq.~[\ref{eq:vw_eff}]) of \begin{align} v_{w}^{\rm Ia} =\sqrt{\frac{2 t_h R_{\rm Ia}
E_{\rm Ia}}{\eta}} \label{eq:vw_sne}.
\end{align} The thermal energy injected by each SN Ia, $E_{\rm Ia} \simeq
\epsilon_{\rm Ia} 10^{51}$ erg, depends on the efficiency $\epsilon_{\rm Ia}$
with which the initial blast wave energy is converted into bulk or turbulent
motion instead of being lost to radiation.  \cite{Thornton+98}
estimate a radiative efficiency $\epsilon_{\rm Ia} \sim 0.1$,
depending weakly on surrounding density, but \citet{Sharma+14} argues
that $\epsilon_{\rm Ia}$ can be considerably higher, $\sim 0.4$, if
the SNe occur in a hot dilute medium, as may characterize the circumnuclear medium.  Hereafter we adopt $\epsilon_{\rm Ia} = 0.4$ as fiducial.

@@

However, even if SN Ia are rare near the stagnation radius, SN Ia may cap the
SMBH accretion rate by periodically blowing gas out of the nucleus of
low mass galaxies.  Between successive SNe, stars release a
gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$
interior to the Ia radius, which is locally gravitationally bound to the SMBH
by an energy $E_{\rm bind} \sim M_{\rm g}\sigma_0^{2}\mathcal{F}^{2}$, where typically $\mathcal{F} 
\lesssim$ few.  From the above definitions we find that 
\be
\frac{E_{\rm Ia}}{E_{\rm bind}} \sim \frac{(v_{w}^{\rm Ia})^{2}}{2\mathcal{F}^{2}\sigma_0^{2}}.  
\label{eq:blowout}
\ee
Hence, for low mass BHs with $\sigma_0 \ll \mathcal{F}v^{\rm Ia}_{w}/2$, SN Ia are capable of dynamically clearing out gas from radii $\sim r_{\rm Ia} \gtrsim r_{\rm s}$.  Even when heating is sufficiently weak that the stagnation radius formally exceeds $r_{\rm Ia}$, the SMBH accretion rate is still potentially limited to a value
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{r_{\rm Ia}}{r_{\rm inf}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}])
for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$
(eq.~[$\ref{eq:eddr_analytic}$]).  Note, however, that for low mass
SMBHs, $\dot{M}_{\rm Ia}$ generally exceeds the maximum accretion rate
for a thermally stable flow $\dot{M}_{\rm TI}$
(eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe
Ia cannot alone prevent cooling instabilities in the nuclei of low
mass galaxies {\bf AG--careful one needs a separate criterion for
  thermal stability when heating is dominated by the Ias}.

>>>>

Type Ia SNe represent a potentially important source of heating, which
unlike core collapse SNe is present even in an evolved stellar
population.  If each SN Ia injects thermal energy $E_{\rm Ia}$ into
the interstellar medium, and the SN rate per stellar mass is given by
$R_{\rm Ia}$, then the resulting volumetric heating rate of $E_{\rm
  Ia}R_{\rm Ia}$ produces an effective heating parameter
(eq.~[\ref{eq:vw_eff}]) of

\begin{align} v_{w}^{\rm Ia} =\sqrt{\frac{2
      t_h R_{\rm Ia} E_{\rm Ia}}{\eta}} \label{eq:vw_sne}.
\end{align}

The thermal energy injected by each SN Ia, $E_{\rm Ia} \simeq
\epsilon_{\rm Ia} 10^{51}$ erg, depends on the efficiency $\epsilon_{\rm Ia}$
with which the initial blast wave energy is converted into bulk or turbulent
motion instead of being lost to radiation.  \cite{Thornton+98}
estimate a radiative efficiency $\epsilon_{\rm Ia} \sim 0.1$,
depending weakly on surrounding density, but \citet{Sharma+14} argues
that $\epsilon_{\rm Ia}$ can be considerably higher, $\sim 0.4$, if
the SNe occur in a hot dilute medium, as may characterize the circumnuclear medium.  Hereafter we adopt $\epsilon_{\rm Ia} = 0.4$ as fiducial.

@@

However, even if SN Ia are rare near the stagnation radius, SN Ia may
cap the SMBH accretion rate by periodically blowing gas out of the
nucleus of low mass galaxies.  Between successive SNe, stars release a
gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$
interior to the Ia radius, which is locally gravitationally bound to
the SMBH by an energy $E_{\rm bind} \sim M_{\rm
  g}\sigma_0^{2}\mathcal{F}^{2}$, where typically $\mathcal{F}
\lesssim$ few.  From the above definitions we find that

\begin{equation}
\frac{E_{\rm Ia}}{E_{\rm bind}} \sim \frac{(v_{w}^{\rm
    Ia})^{2}}{2\mathcal{F}^{2}\sigma_0^{2}}.
\label{eq:blowout}
\end{equation}

Hence, for low mass BHs with $\sigma_0 \ll \mathcal{F}v^{\rm
  Ia}_{w}/2$, SN Ia are capable of dynamically clearing out gas from
radii $\sim r_{\rm Ia} \gtrsim r_{\rm s}$.  Even when heating is
sufficiently weak that the stagnation radius formally exceeds $r_{\rm
  Ia}$, the SMBH accretion rate is still potentially limited to a
value
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{r_{\rm Ia}}{r_{\rm inf}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}])
for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$
(eq.~[$\ref{eq:eddr_analytic}$]).  Note, however, that for low mass
SMBHs, $\dot{M}_{\rm Ia}$ generally exceeds the maximum accretion rate
for a thermally stable flow $\dot{M}_{\rm TI}$
(eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe
Ia cannot alone prevent cooling instabilities in the nuclei of low
mass galaxies.
